For Induction-Induction, Induction is Enough

### 2019-11-12 paper/universaliit.tex
finished related work
small typo in sec 3.
grammar and typos

Note that $(\Piu \,A\,B)[\sigma]$ should be defined as $\Piu
\,(A[\sigma])\,(B[\varu\, \mathsf{0} \,\consu\, \wkO\,\sigma])$, and thus we need to define
$\wkO$, the weakening of substitutions.  The basic idea is to increment the de
Bruijn indices of all the variables.  Actually this is not so simple because of
the $\Piu$ type: indeed, we want to define $\wkO\,(\Piu\,A\,B)$ as the $\Pi$ type
of the weakening of $A$ and $B$, but here, $B$ must be weakened with respect to
the second last variable of the context, rather than the last one.  For this
reason, we need to generalize the weakening as occuring anywhere in the context.
\begin{alignat*}{5}
  & \wk_n && :  \Tyu\ra\Tyu \\
  & \wk_n && :  \Tmu\ra\Tmu \\
  % & \wk && : \N \ra \N\ra\N \\
  & \wkO && : \Subu\ra \Subu
  \end{alignat*}
  The natural number $n$ specifies at which position of the context the
  weakening occurs.
  Here, $\wkO$ weakens with respect to the last variable.

Later, in Section~\ref{ss:weaken-typing}, we show that weakening preserves
typing. Stating a typing rule for this operation requires to have a way to
express the weakening occuring at the middle of a context. We consider pairs of
untyped contexts, which should be thought of as a splitting of a context at some
position. The full context is recovered by merging the two components:
% (which appears in the typing rule of a weakening occuring in the middle of a context),
% and the length $\length{\Gamma}$ of a
% context $\Gamma$:
\begin{alignat*}{5}
  & \blank \merge \blank && :  \Conu\ra\Conu\ra\Conu \\
  & \Gamma \merge \cdot && :=  \Gamma \\
  & \Gamma \merge (\Delta \extu A) && :=  (\Gamma \merge \Delta)\extu A
\end{alignat*}
We think of the second context as a telescope over the first one. We also define
weakening for telescopes, which will be used to give typing rules for telescopes
in Section \ref{sec:typing_judgments}:
\begin{alignat*}{5}
  & \wkO && : \Conu\ra \Conu
  \\
  & \wkO\,\emptyu && := \emptyu
  \\
  & \wkO\,(\Delta\,\extu A) && := \wkO\,\Delta\,\extu \wk_{\length{\Delta}}\,A
  \end{alignat*}
$\length{\Gamma}$ denotes the length of a the context $\Gamma$.

@@

\subsection{Relating the Syntax to a Model}
It remains to show that the constructed syntactic model is initial. To achieve
this, we first define a relation between the syntactic model and an arbitrary
model, then show that the relation is functional, which lets us extract a
homomorphism from the relation.

@@

\subsection{Left-Totality and the Recursor}
For the recursor, we build a morphism from the syntactic model to the semantic
one. The image of a syntactic context is a unique semantic context which is
related to it, and similarly for types and terms.
% We construct it as the semantic
% Induction on the typing judgments shows that any QIIT morphism from the
% syntactic model $S$  to a semantic model $M$ sends well-typed syntax on a related
% semantic counter-part.
Thus, as a first step, we use induction on well-formedness judgments to construct
semantic counterparts:
  \begin{alignat*}{5}
     & \Sigma\Con{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    \sum_{\model{\Gamma}} \wf{\Gamma} \sim \model{\Gamma}
    \\
     & \Sigma\Ty{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{A} : \Gamma \vdash A ) \ra
    ({\model{A}:\model{\Ty}\model\Gamma})\times ( \wf{A} \sim \model{A})
    \\
     & \Sigma\Tm{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{A} : \Gamma \vdash A ) \ra
    ( \wf{A} \sim \model{A}) \ra
    \\ & && \qquad
    (\wf{t} : \Gamma \vdash t \in A ) \ra
    ({\model{t}:\model{\Tm}\model\Gamma}\model{A})\times ( \wf{t} \sim \model{t})
    \\
     & \Sigma\Var{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{A} : \Gamma \vdash A ) \ra
    ( \wf{A} \sim \model{A}) \ra
    \\ & && \qquad
    (\wf{x} : \Gamma \vdash x \invar A ) \ra
    ({\model{x}:\model{\Tm}\model\Gamma}\model{A})\times ( \wf{x} \sim \model{x})
    \\
     & \Sigma\Sub{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{\Delta} : \Delta \vdash  ) \ra
    ( \wf{\Delta} \sim \model{\Delta}) \ra
    \\ & && \qquad
    (\wf{\sigma} : \Gamma \vdash \sigma \Rightarrow \Delta ) \ra
    ({\model{\sigma}:\model{\Sub}\model\Gamma}\model{\Delta})\times ( \wf{\sigma} \sim \model{\sigma})
  \end{alignat*}
  The right uniqueness of the relation is used in this induction. It is then
straightforward to show (without induction, but using right uniqueness) that the
first projection of these constructions yield a model morphism from the syntax
to the model.

>>>>

Note that $(\Piu \,A\,B)[\sigma]$ should be defined as $\Piu
\,(A[\sigma])\,(B[\wkO\,\sigma \,\consu\, \varu\,\mathsf{0}])$, and thus we need to define
$\wkO$, the weakening of substitutions.  The basic idea is to increment the de
Bruijn indices of all the variables.  Actually, this is not so simple because of
the $\Piu$ type: we want to define $\wkO\,(\Piu\,A\,B)$ as the $\Pi$ type
of the weakening of $A$ and $B$, but here, $B$ must be weakened with respect to
the second last variable of the context, rather than the last one.  For this
reason, we need to generalise the weakening as occuring anywhere in the context.
\begin{alignat*}{5}
  & \wk_n && :  \Tyu\ra\Tyu \\
  & \wk_n && :  \Tmu\ra\Tmu \\
  % & \wk && : \N \ra \N\ra\N \\
  & \wkO && : \Subu\ra \Subu
  \end{alignat*}
  The natural number $n$ specifies at which position of the context the
  weakening occurs.
  Here, $\wkO$ weakens with respect to the last variable.

Later, in Section~\ref{ss:weaken-typing}, we show that weakening preserves
typing. Stating a typing rule for this operation requires weakening at the
middle of a context. We consider pairs of untyped contexts, which should be
thought of as a splitting of a context at some position. The full context is
recovered by merging the two components:
% (which appears in the typing rule of a weakening occuring in the middle of a context),
% and the length $\length{\Gamma}$ of a
% context $\Gamma$:
\begin{alignat*}{5}
  & \blank \merge \blank && :  \Conu\ra\Conu\ra\Conu \\
  & \Gamma \merge \cdot && :=  \Gamma \\
  & \Gamma \merge (\Delta \extu A) && :=  (\Gamma \merge \Delta)\extu A
\end{alignat*}
We think of the second context as a telescope over the first one. We also define
weakening for telescopes, which will be used to give typing rules for telescopes
in Section \ref{sec:typing_judgments}:
\begin{alignat*}{5}
  & \wkO && : \Conu\ra \Conu
  \\
  & \wkO\,\emptyu && := \emptyu
  \\
  & \wkO\,(\Delta\,\extu A) && := \wkO\,\Delta\,\extu \wk_{\length{\Delta}}\,A
  \end{alignat*}
$\length{\Gamma}$ denotes the length of a the context $\Gamma$.

@@

\subsection{Relating the Syntax to a Model}
\label{sec:streicher}
It remains to show that the constructed syntactic model is initial. To achieve
this, we first define a relation between the syntactic model and an arbitrary
model, then show that the relation is functional, which lets us extract a
homomorphism from the relation.

@@

\subsection{Left-Totality and the Recursor}
For the recursor, we build a morphism from the syntactic model to the semantic
one. The image of a syntactic context is a unique semantic context which is
related to it, and similarly for types and terms.
% We construct it as the semantic
% Induction on the typing judgments shows that any QIIT morphism from the
% syntactic model $S$  to a semantic model $M$ sends well-typed syntax on a related
% semantic counter-part.
Thus, as a first step, we use induction on well-formedness judgments to construct
semantic counterparts:
  \begin{alignat*}{5}
     & \Sigma\Con{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    \sum_{\model{\Gamma}} \wf{\Gamma} \sim \model{\Gamma}
    \\
     & \Sigma\Ty{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{A} : \Gamma \vdash A ) \ra
    ({\model{A}:\model{\Ty}\model\Gamma})\times ( \wf{A} \sim \model{A})
    \\
     & \Sigma\Tm{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{A} : \Gamma \vdash A ) \ra
    ( \wf{A} \sim \model{A}) \ra
    \\ & && \qquad
    (\wf{t} : \Gamma \vdash t \in A ) \ra
    ({\model{t}:\model{\Tm}\model\Gamma}\model{A})\times ( \wf{t} \sim \model{t})
    \\
     & \Sigma\Var{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{A} : \Gamma \vdash A ) \ra
    ( \wf{A} \sim \model{A}) \ra
    \\ & && \qquad
    (\wf{x} : \Gamma \vdash x \invar A ) \ra
    ({\model{x}:\model{\Tm}\model\Gamma}\model{A})\times ( \wf{x} \sim \model{x})
    \\
     & \Sigma\Sub{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{\Delta} : \Delta \vdash  ) \ra
    ( \wf{\Delta} \sim \model{\Delta}) \ra
    \\ & && \qquad
    (\wf{\sigma} : \Gamma \vdash \sigma \Rightarrow \Delta ) \ra
    ({\model{\sigma}:\model{\Sub}\model\Gamma}\model{\Delta})\times ( \wf{\sigma} \sim \model{\sigma})
  \end{alignat*}
  The right uniqueness of the relation is used in this induction. It is then
straightforward to show (without induction, but using right uniqueness) that the
first projection of these constructions yields a model morphism from the syntax
to the model.
